---
title: "Instructor Ratings with Multiple Linear Regression"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


# Executive Summary

**Objectives & Problem Statement**

This analysis investigates the primary factors influencing student evaluations of their instructors, aiming to isolate which attributes, course characteristics, and demographic variables drive perceived quality. By using rigorous statistical modeling, this project quantifies these relationships to separate meaningful indications from noise and evaluate potential structural inequalities in rating metrics.

**Data & Methodology**

The analysis was conducted using a comprehensive dataset tracking instructor ratings alongside a diverse group of course and rater metrics. The analytical workflow, built entirely within R and the Tidyverse ecosystem involved:

-  **Exploratory Data Analysis (EDA):**Assessing multicollinearity through correlation matrices to understand overlapping variance among rating metrics.

-  **Model Progression:**Fitting an initial 16-variable global additive regression model, executing stepwise variable selection, introducing complex interaction terms, and finalizing a variance-stabilizing response transformation.

-  **Diagnostic Validation:**Testing Gauss-Markov assumptions using both visual and statistical diagnostics. This included analyzing **Residuals vs Fitted plots** to check for constant variance (homoscedasticity) and **Normal Q-Q plots** to evaluate normality, backed by Breusch-Pagan and Shapiro-Wilk for residual normality to ensure inferential validity.

**Key Findings & Insights**

-  **Primary Drivers:**Core instructional traits, specifically student-perceived easiness ($p < 0.001$), clarity ($p < 0.001$), and helpfulness ($p < 0.001$) emerged as the most statistically significant positive drivers of overall instructor quality ratings.

-  **Interaction Effects:** The impact of individual variables is heavily moderated by context. Notably, interaction terms involving course workload (numCourses) and variance in student interest (sdRaterInterest) revealed critical shifts in how evaluations scale.

-  **Statistical Rigor:** While early additive models violated standard assumptions regarding residual variance and normality, the final model successfully achieved a stabilized residual distribution ($p_{\text{BP}} = 0.1198$, $p_{\text{Shapiro}} = 0.1586$) through a power transformation of the response variable, guaranteeing that the final coefficient estimates are both efficient and unbiased.

# Data Preparation & Setup

### Libraries

```{r}
#Core data manipulation and visualization (loads dplyr, ggplot2, tidyr, etc.)
library(tidyverse)
library(here)

#Modeling, diagnostics, and tidying
library(car)
library(MASS)
library(faraway)
library(lmtest)
library(broom)

#Formatting and reporting
library(knitr)
library(kableExtra)
library(bookdown)
library(ggcorrplot)
library(patchwork)
library(ggfortify)
```

### Helper Functions

```{r}
#Reusable function to build standard vertical regression statistics tables
create_model_summary <- function(model_object, table_caption) {
  g_stats <- glance(model_object)
  
  stat_df <- data.frame(
    " " = c("F - Statistic", 
            "Residual Standard Error", 
            "Multiple R - Squared", 
            "Adjusted R - Squared"),
    "Value" = c(
      round(g_stats$statistic, 4),
      round(g_stats$sigma, 4),
      round(g_stats$r.squared, 4),
      round(g_stats$adj.r.squared, 4)
    ),
    "DF" = c(
      paste0(g_stats$df - 1, " and ", g_stats$df.residual),
      as.character(g_stats$df.residual),
      NA,
      NA
    ),
    "Pr(>|t|)" = c(
      format.pval(g_stats$p.value, digits = 4),
      NA,
      NA,
      NA
    ),
    check.names = FALSE
  )
  
  options(knitr.kable.NA = ' ')
  
 kable(
    stat_df,
    caption = table_caption,
    align = c("l", "c", "c", "c"),
    format = "html"
  ) |>
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"), 
      full_width = FALSE, 
      position = "center"
    )
}
```


### Load Data

```{r}
library(alr4) 
data("Rateprof")
```

# Exploratory Data Analysis

### Preliminary Exploratory Analysis: Pairs Plot

```{r}
#Define a smooth panel function to add trend lines
panel.smooth.custom <- function (x, y, ...) {
    points(x, y, col = rgb(0, 0, 0, 0.25), pch = 16, cex = 0.6) # Semi-transparent points
    lines(lowess(x, y), col = "firebrick", lwd = 2)             # Smooth red trend line
}

#Run the pairs plot using your dataset columns
pairs(Rateprof[, c("quality", "helpfulness", "clarity", "easiness", 
                          "raterInterest", "numYears", "numRaters", "numCourses")],
      lower.panel = panel.smooth.custom,   # Applies the points and lines to the lower triangle
      upper.panel = panel.smooth.custom,   # Applies to the upper triangle (or set to NULL to reduce clutter)
      main = "Preliminary Scatterplot Matrix of Instructor Metrics",
      gap = 0.4,                           # Tightens the spacing between panels
      labels = c("Quality", "Helpfulness", "Clarity", "Easiness", 
                 "Rater Interest", "Num Years", "Num Raters", "Num Courses"))
```

### Correlation Matrix

```{r heatmap, fig.width=10, fig.height=8, fig.align='center'}

#Create data frame to convert variables to numeric to use for cor()
num_df <- Rateprof
#Convert to numeric
num_df <- num_df %>%
  mutate(across(where(is.factor), as.numeric))

#Compute correlation matrix
cor_matrix <- cor(num_df)

#Plot an elegant heatmap showing only the lower triangle
p_corr <- ggcorrplot(cor_matrix, 
           hc.order = TRUE, 
           type = "lower",
           lab = TRUE,           # Adds text labels inside squares
           lab_size = 3,         # Perfect size for many variables
           method = "square", 
           colors = c("#6D9EC1", "white", "#E46726"), # Clean professional gradient
           title = "Correlation Heatmap of Model Variables",
           ggtheme = theme_minimal())
p_corr
#Save as PNG
ggsave(
  filename = "../figures/correlation_heatmap.png", 
  plot = p_corr, 
  width = 8, 
  height = 6, 
  dpi = 300
)
```

# Methodology

### Population Regression Equation

To investigate the driving factors of instructor ratings, we first specify the global Population Regression Equation. This theoretical model incorporates all 16 main-effect predictors before any variable selection or interaction transformations are applied:

$$
\begin{aligned}
\widehat{\text{quality}} ={} & \hat{\beta}_0 + \hat{\beta}_1\text{gender} + \hat{\beta}_2\text{numYears} + \hat{\beta}_3\text{numRaters} + \hat{\beta}_4\text{numCourses} \\
& + \hat{\beta}_5\text{pepper} + \hat{\beta}_6\text{discipline} + \hat{\beta}_7\text{dept} + \hat{\beta}_8\text{helpfulness} \\
& + \hat{\beta}_9\text{clarity} + \hat{\beta}_{10}\text{easiness} + \hat{\beta}_{11}\text{raterInterest} + \hat{\beta}_{12}\text{sdQuality} \\
& + \hat{\beta}_{13}\text{sdHelpfulness} + \hat{\beta}_{14}\text{sdClarity} + \hat{\beta}_{15}\text{sdEasiness} + \hat{\beta}_{16}\text{sdRaterInterest}
\end{aligned}
$$

Using our sample dataset, we will fit this model via Ordinary Least Squares (OLS) to derive our Estimated Regression Equation, allowing us to analyze individual predictor significance and test for Gauss-Markov assumptions.

## Model Fitting

### Fit the Model: Full Additive Model

```{r}
#Full additive model
lm1 <- lm(quality ~ gender + numYears + numRaters 
    + numCourses + pepper + discipline + dept  
    + helpfulness + clarity + easiness   
    + raterInterest + sdQuality + sdHelpfulness
    + sdClarity + sdEasiness 
    + sdRaterInterest, Rateprof)
```

[Summary output for full additive model.]{.underline}

```{r, results='asis'}

#Model I Summary Statistics Table
#For Model I
create_model_summary(lm1, "Regression Statistics: Full Additive Model I")
```

### Check Assumptions

[Constant Variance and Normality]{.underline}

A residuals versus fitted values plot with Breusch - Pagan test. And a normal Q - Q plot with Shapiro - Wilk test.

```{r model-1-diagnostics, fig.show = "hold", out.width = "50%", dpi = 100, fig.cap = "Diagnostic Plots for the Full Additive Model I"}

#Augment the model to get a tidy data frame with residuals (.resid) & fitted values (.fitted)
diag_df <- augment(lm1)

#Residuals vs Fitted Plot
p_res <- ggplot(diag_df, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(se = FALSE, color = "blue", method = "loess", formula = y ~ x) +
  labs(
    title = "Baseline Model: Residuals vs Fitted",
    x = "Fitted Values",
    y = "Residuals"
  ) +
  theme_minimal(base_size = 12)

#Normal Q-Q Plot
p_qq <- ggplot(diag_df, aes(sample = .resid)) +
  stat_qq(alpha = 0.6) +
  stat_qq_line(color = "red") +
  labs(
    title = "Baseline Model: Normal Q-Q",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_minimal(base_size = 12)

#Combine side-by-side & save cleanly with ggsave()
baseline_diag <- p_res + p_qq
baseline_diag
ggsave(
  filename = "../figures/baseline_diagnostics.png",
  plot = baseline_diag,
  width = 10,
  height = 4.5,
  dpi = 100
)
#Breusch- Pagan Test
bp1 <- bptest(lm1)
bp1

#Shapiro - Wilk Test
shap1 <- shapiro.test(resid(lm1))
shap1

```

### Fit the model: Use the stepwise function on the full additive model.

```{r}
#Recall:
lm1 <- lm(quality ~ gender + numYears + numRaters 
    + numCourses + pepper + discipline + dept + 
    + helpfulness + clarity + easiness +  raterInterest
    + sdQuality + sdHelpfulness + sdClarity + sdEasiness 
    + sdRaterInterest, Rateprof)

#Step() with full additive model
lm2 <- step(lm1, trace = 0)
```

[Summary output for the stepwise model from the full additive model.]{.underline}

```{r, results='asis'}

#Model II Summary Statistics
create_model_summary(lm2, "Regression Statistics: Stepwise Additive Model II")
```

### Check Assumptions

[Constant Variance and Normality]{.underline}

A residuals versus fitted values plot with Breusch - Pagan test. And a normal Q - Q plot with Shapiro - Wilk test.

```{r model-2-diagnostics, fig.show = "hold", out.width = "50%", dpi = 100, fig.cap = "Diagnostic Plots for the Stepwise Additive Model II"}
#Residuals vs Fitted
plot(lm2, which = 1, col = rgb(0, 0, 0, 0.6), pch = 16)

#Normal Q-Q Plot
plot(lm2, which = 2, col = rgb(0, 0, 0, 0.6), pch = 16)

save.image()

#Breusch- Pagan Test
bp2 <- bptest(lm2)
bp2

#Shapiro - Wilk Test
shap2 <- shapiro.test(resid(lm2))
shap2
```

### Fit the model: Interactions model

Remove helpfulness and clarity due to potential strong linear relationship between the two predictors and the response variable quality assessed by the pairs plot.

```{r}
#Interactions model introduced
#Full model -helpfulness, -clarity

lm3 <- lm(quality ~  discipline + dept + 
      (gender + pepper + easiness + 
      raterInterest + numYears + numCourses  
      + numRaters + sdHelpfulness + sdClarity + 
      sdEasiness + sdRaterInterest)^2, Rateprof)

```


[Summary output for the interactions model.]{.underline}

```{r, results='asis'}

#Model III Summary Statistics
create_model_summary(lm3, "Regression Statistics: Stepwise Interactions Model III")
```

### Check Assumptions

[Constant Variance and Normality]{.underline}

A residuals versus fitted values plot with Breusch - Pagan test. And a normal Q - Q plot with Shapiro - Wilk test.

```{r model-3-diagnostics, fig.show = "hold", out.width = "50%", dpi = 100, fig.cap = "Diagnostic Plots for the Stepwise Interactoins Model III"}
#Residuals vs Fitted
plot(lm3, which = 1, col = rgb(0, 0, 0, 0.6), pch = 16)

#Normal Q-Q Plot
plot(lm3, which = 2, col = rgb(0, 0, 0, 0.6), pch = 16)

save.image()

#Breusch- Pagan Test
bp3 <- bptest(lm3)
bp3

#Shapiro - Wilk Test
shap3 <- shapiro.test(resid(lm3))
shap3
```

### Transformed Model

The assumptions are still violated, therefore check for transformations.

```{r}
#Extract boxcox coordinates into a data frame (plot = FALSE prevents base R from plotting)
bc_data <- boxcox(lm3, lambda = seq(2.2, 5.0, by = 0.05), plot = FALSE)
bc_df <- data.frame(lambda = bc_data$x, loglik = bc_data$y)

#Find optimal lambda (maximum log-likelihood)
opt_lambda <- bc_df$lambda[which.max(bc_df$loglik)]

#Build ggplot
p_boxcox <- ggplot(bc_df, aes(x = lambda, y = loglik)) +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_vline(xintercept = opt_lambda, linetype = "dashed", color = "red") +
  annotate(
    "text", 
    x = opt_lambda + 0.15, 
    y = min(bc_df$loglik) + 2, 
    label = paste0("Optimal λ = ", round(opt_lambda, 2)), 
    color = "red", 
    hjust = 0
  ) +
  labs(
    title = "Box-Cox Log-Likelihood Profile",
    subtitle = "Transforming response variable for Model III",
    x = expression(lambda),
    y = "Log-Likelihood"
  ) +
  theme_minimal(base_size = 12)
p_boxcox
#Save cleanly via ggsave
ggsave(
  filename = here("figures", "boxcox_transformation.png"),
  plot = p_boxcox,
  width = 8,
  height = 5,
  dpi = 300
)
```

# Results

[Final Model: Stepwise Transformed model]{.underline}

Model has been transformed and the stepwise function has been applied to the transformed model.

```{r}

#Transformed model
lm4 <- lm((quality^3.55) ~  discipline + dept + 
    (gender + pepper + easiness +  raterInterest 
    + numYears + numCourses  + numRaters + sdQuality 
    + sdHelpfulness + sdClarity + sdEasiness 
    + sdRaterInterest)^2, Rateprof)

#Apply step function
lm5 <- step(lm4, trace = 0)
```

[Summary output for final model.]{.underline}

```{r, results='asis'}

#Tidy the model coefficients dynamically
tidy_coef <- tidy(lm5) |>
  mutate(
    across(where(is.numeric), ~ round(., 4)),
    # Add standard scientific significance indicator strings
    Significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      p.value < 0.1   ~ ".",
      TRUE            ~ ""
    )
  )

#Build a beautiful, structured table
kable(
  tidy_coef,
  col.names = c("Predictor", "Estimate", "Std. Error", "t statistic", "Pr(>|t|)", ""),
  caption = "Regression Coefficients: Interactions Model IV",
  format = "html"  # Changed to HTML
) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), 
    full_width = TRUE
  ) |>
  row_spec(which(tidy_coef$p.value < 0.05), bold = TRUE, background = "#f9f9f9") %>%
  pack_rows("Main Effects", 1, 13) |>
  pack_rows("Interaction Terms", 14, nrow(tidy_coef))

```
[Summary output for full additive model.]{.underline}

```{r, results='asis'}
#Model IV Summary Statistics
create_model_summary(lm5, "Regression Statistics: Transformed Model IV")
```

### Check Assumptions

[Constant Variance and Normality]{.underline}

A residuals versus fitted values plot with Breusch - Pagan test. And a normal Q - Q plot with Shapiro - Wilk test.

```{r model-4-diagnostics, fig.show = "hold", out.width = "50%", dpi = 100, fig.cap = "Diagnostic Plots for the Final Transformed Model IV"}

#Augment the model to get a tidy data frame with residuals (.resid) & fitted values (.fitted)
diag_df_final <- augment(lm5)

#Residuals vs Fitted Plot
p_res_final <- ggplot(diag_df_final, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(se = FALSE, color = "blue", method = "loess", formula = y ~ x) +
  labs(
    title = "Final Model: Residuals vs Fitted",
    x = "Fitted Values",
    y = "Residuals"
  ) +
  theme_minimal(base_size = 12)

#Normal Q-Q Plot
p_qq_final <- ggplot(diag_df_final, aes(sample = .resid)) +
  stat_qq(alpha = 0.6) +
  stat_qq_line(color = "red") +
  labs(
    title = "Final Model: Normal Q-Q",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_minimal(base_size = 12)
#Combine side-by-side & save cleanly with ggsave()
final_diagnostics <- p_res_final + p_qq_final
final_diagnostics

ggsave(
  filename = "../figures/final_diagnostics.png",
  plot = final_diagnostics,
  width = 10,
  height = 4.5,
  dpi = 100
)
#Breusch- Pagan Test
bp5 <- bptest(lm5)
bp5

#Shapiro - Wilk Test
shap5 <- shapiro.test(resid(lm5))
shap5
```

### Conclusions

To address multicollinearity and remove non-significant higher-order terms, stepwise AIC selection was performed on the transformed interaction model ($\text{Model IV}$). The resulting parsimonious model retains key main effects along with 23 moderating two-way interaction terms that best optimize predictive performance and model fit:

```{r}
#Final Model
# Display the final model selected by stepwise regression
formula(lm5)
```
### Comparison of Full Additive Model I and Final Model

```{r}
# 3. Combine into a 2x2 grid (Top row: Baseline, Bottom row: Final)
diagnostic_comparison <- (p_res + p_qq) / (p_res_final + p_qq_final)

# 4. Save using ggsave()
ggsave(
  filename = "../figures/model_diagnostics_comparison.png",
  plot = diagnostic_comparison,
  width = 10,
  height = 8,
  dpi = 300
)
```

### Table: Comparison of All Models

The table includes: adjusted $R^2$, Shapiro - Wilk p - value, and Breusch - Pagan p - value.

The models: full additive model I, full stepwise model II, interactions model III, and final model IV.

```{r}
#Dynamically extract the p-values from your saved test objects
models_tb <- data.frame(
  "Model" = c("Full Additive I", "Full Step II", "Interactions III", "Final IV"),
  
  "Adj. $R^2$" = c(
    round(glance(lm1)$adj.r.squared, 4),
    round(glance(lm2)$adj.r.squared, 4),
    round(glance(lm3)$adj.r.squared, 4),
    round(glance(lm5)$adj.r.squared, 4)
  ),
  
  "Shapiro - Wilk" = c(
    format.pval(shap1$p.value, digits = 4),
    format.pval(shap2$p.value, digits = 4),
    format.pval(shap3$p.value, digits = 4),
    format.pval(shap5$p.value, digits = 4)
  ),
  
  "Breusch - Pagan" = c(
    format.pval(bp1$p.value, digits = 4),
    format.pval(bp2$p.value, digits = 4),
    format.pval(bp3$p.value, digits = 4),
    format.pval(bp5$p.value, digits = 4)
  ),
  
  check.names = FALSE
)

kable(
  models_tb,
  caption = "Regression Models Comparison",
  align = c("l", "c", "c", "c"),
  format = "html"  # Changed to HTML
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), 
    full_width = FALSE, 
    position = "center"
  )

```
